Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2022-11-23 text/Manuscript.Rmd
MS -editing results section until NIT-GC immunos

---
title: Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: "*Kei Jokura, Nobuo Ueda, Martin Gühmann, Luis Alfonso Yañez-Guerra, Kyle C. A. Wedgwood, Piotr Słowiński, Gáspár Jékely*"
date: "*Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4
  4QD, UK*"
output:
  pdf_document:
    latex_engine: lualatex
    template: default
  html_document:
    theme: lumen
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology. In nervous systems, NO can modulate circuits, but its effects on neural dynamics and behaviour remain unclear. Here we show how NO signalling shapes circuit dynamics to orchestrate light-avoidance behaviour. We studied UV-avoidance mediated by brain ciliary photoreceptor cells (cPRCs) in larval *Platynereis dumerilii*, a marine annelid. We found NOS specifically expressed in interneurons (INNOS) postsynaptic to cPRCs. Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production. NO feeds back to cPRCs and triggers delayed cPRC activation through an unconventional guanylyl cyclase. This results in the activation of a second population of postsynaptic interneurons (INRGW). It was also suggested that activation of INRGWs activates serotonergic ciliomotor neurons (Ser-h1), which are postsynaptic and involved in increasing ciliary beat frequency. In *NOS* knockout larvae, NO feedback and INRGW activation do not occur and avoidance behaviour is defective. Our results reveal how NO-mediated neuroendocrine signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In 2 and 3-day-old larvae, we detected *NOS* expression in two or four cells in the apical organ region (Figure 1A-D). The same four cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene in 3-day-old larvae(Figure 1E). This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the 3-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1F and G). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1H).

@@

**Nitric oxide signalling mediates UV-avoidance behaviour** The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 2A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation [@veraszto2018]. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 2B and C). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@veraszto2018].

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans [@Moroz2020]. To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Figure 3---figure supplement 1). Using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017] to find candidate NIT-GCs involved in the cPRC circuit among 15 NIT-GCs, two NIT-GCs (NIT-GC1 and NIT-GC2) were found to be highly expressed in the cPRC (Figure 3---figure supplement 1). Expression analysis with c-opsin1 and MLD pNPs (pedal peptide neuropeptide precursor 2) [@Arendt2004; @williams2017], which have already been identified as cPRC markers using in situ HCR, showed that both NIT-GC1 and NIT-GC2 are strongly and specifically expressed in cPRC (Figure 3A and B, Figure 3---figure supplement 2). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Figure 3C and D). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). These two localisation differences suggest the existence of different intracellular signalling mechanisms in the ciliary and neuroaxonal regions within the cPRC.

>>>>

---
title: Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance
  in zooplankton
author: "*Kei Jokura, Nobuo Ueda, Martin Gühmann, Luis Alfonso Yañez-Guerra, Piotr Słowiński, Kyle C. A. Wedgwood, Gáspár Jékely*"
date: "*Living Systems Institute, University of Exeter, Stocker road, Exeter, EX4
  4QD, UK*"
output:
  pdf_document:
    latex_engine: lualatex
    template: default
  html_document:
    theme: lumen
  word_document: default
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
---

@@

Nitric oxide (NO) generated by nitric-oxide synthase (NOS) is a key regulator of animal physiology. Here we uncover a role for NO in shaping circuit dynamics to orchestrate light-avoidance behaviour. We studied UV-avoidance mediated by brain ciliary photoreceptors (cPRCs) in larval *Platynereis dumerilii*, a marine annelid. We found NOS specifically expressed in interneurons (INNOS) postsynaptic to cPRCs. Stimulation of cPRCs by violet light leads to cPRC inhibition but concomitant INNOS activation and NO production. NO feeds back to cPRCs and triggers their delayed activation through an unconventional guanylyl cyclase. This results in the activation of projection interneurons and the inhibition of serotonergic ciliomotor neurons. In *NOS* mutants, NO feedback and projection-neuron activation do not occur and avoidance behaviour is defective. By mathematical modelling, we could capture the circuit dynamics in both wild-type and mutant larvae. Our results reveal how NO-mediated neuroendocrine signalling gates a synaptic circuit to initiate light-avoidance behaviour.

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two- and three-day-old larvae, we detected *NOS* expression in four cells (two of them weakly expressing) in the apical organ region (Figure 1A-D). The same four cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene (Figure 1E). This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1F and G). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1C, D). INRGW neurons synapse on the head serotonergic ciliomotor neurons (Ser-h1), which synapse on the prototroch ciliary band and the cholinergic MC ciliomotor neurons (Figure 1C, D). We also detected *NOS* expression by HCR in the region of the visual eyes and the pigmented eyespots.

@@

**Nitric oxide signalling mediates UV-avoidance behaviour**

The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 2A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm violet light. We also tested phototaxis, by exposing larvae to 480 nm directional and collimated light from the top of the column. As previously shown, wild-type larvae swim downward within 30 sec following non-directional UV/violet light stimulation [@veraszto2018]. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 2B and C). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@veraszto2018]. *NOS*-mutant larvae also showed reduced phototactic behaviour, suggesting a function in the visual eyes mediating phototaxis [@Randel2014a], where *NOS* is also expressed (Figure 1). To test if the observed defects are due to a role of NOS in neuronal signalling or circuit developmental, we also tested larvae exposed to a NOS inhibitor, L-NAME. Larvae incubated for xy min in 0.1 mM or 1 mM L-NAME showed defective UV avoidance, but normal phototaxis. Overall, our results indicate an acute requirement for NOS signalling in UV-avoidance and a possible indirect, developmental role in the visual system.

@@

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit ([@Veraszto2017] and data not shown). Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in _Platynereis_, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and [**how many?] choanoflagellate. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 _Platynereis_ sequences were parts of several major and ancient clades of the tree (Figure 4---figure supplement 1). 
To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, four showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 3---figure supplement 1). In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers (Figure 3A and B, Figure 3---figure supplement 2). The INNOS cells were identified by NOS expression and their spatial mapping. 
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out HCR with probes for _NIT-GC1_ and _NIT-GC2_ mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated tubulin antibody and with an HCR probe against _MLD proneuropeptide_. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs (Figure 3C and D). These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.
